\documentclass[10pt]{article}
\input{../../notes-style.tex}

\usepackage{multicol}
\usepackage{array}

\title{Ch.~11 Notes\\\large B-trees}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-2em}

\begin{multicols}{2}

\subsection*{Why B-trees}
Binary BSTs are memory-bound on pointer chases -- each hop is a
cache miss. A B-tree stuffs many keys per node so each disk / cache
block read pays off: height becomes $O(\log_m n)$ for order $m$.
The standard file-system / database index.

\subsection*{Definition (order $m$)}
For every non-root internal node:
\begin{itemize}
    \item Has $k$ keys, $\lceil m/2 \rceil - 1 \le k \le m - 1$.
    \item Has $k+1$ children.
    \item Keys within a node are sorted; child $i$ holds everything
    between key $i-1$ and key $i$.
    \item All leaves at the \emph{same depth}.
\end{itemize}
Root relaxed: may have as few as 1 key. \textbf{2-3-4 tree} = order 4
(keys 1--3, children 2--4).

\subsection*{Node skeleton}
\begin{lstlisting}[language=C++]
struct BTNode {
    bool isLeaf;
    std::vector<int>       keys;      // up to m-1
    std::vector<BTNode*>   children;  // up to m
};
\end{lstlisting}
Node arrays kept sorted; binary-search within a node.

\subsection*{Height}
Fully packed: $n \approx m^{h+1} \Rightarrow h = O(\log_m n)$.
For $m = 100$ and $n = 10^9$: $h \approx 5$.

\subsection*{Search}
Within a node, binary-search for the key or the child slot.
\begin{lstlisting}[language=C++]
int i = 0;
while (i < node->keys.size() && key > node->keys[i]) ++i;
if (i < node->keys.size() && key == node->keys[i]) return hit;
if (node->isLeaf) return miss;
return search(node->children[i], key);
\end{lstlisting}
$O(m \log_m n) = O(\log n)$, but every ``step'' is one block read.

\subsection*{Split (the core insert primitive)}
Node with $m-1$ keys (full) needs to grow:
\begin{enumerate}
    \item Take the \emph{median} key; push it up into the parent.
    \item Left sibling gets the left half; new right sibling gets the
    right half. Children split accordingly.
    \item If parent is now full, split it too -- cascades up.
\end{enumerate}
Root split creates a new root $\to$ tree grows by one level
(the only way B-trees grow).

\columnbreak

\subsection*{Insert (preemptive split)}
Descend toward the leaf; whenever you're about to enter a full node,
split it first. Then insert into the leaf -- guaranteed to have room.
\begin{itemize}
    \item One downward pass; no retrace.
    \item $O(\log n)$ node reads, $O(1)$ splits per level on average.
\end{itemize}

\subsection*{Rotation (sibling borrow)}
Transfers one key to an under-full node from a sibling through the
parent, preserving sort order.
\begin{itemize}
    \item \textbf{Right-rotate on node}: borrow from left sibling via
    parent.
    \item \textbf{Left-rotate on node}: borrow from right sibling.
\end{itemize}
Both are $O(m)$.

\subsection*{Fusion (merge) -- the remove primitive}
Under-full node + under-full sibling $\Rightarrow$ pull the separator key
down from the parent and concatenate the two into one node.
If the parent is now under-full, recurse upward. Root fusion
collapses a level -- the only way B-trees shrink.

\subsection*{Remove (preemptive merge)}
Descending to the key:
\begin{itemize}
    \item Before entering an under-full child, \textbf{rotate} from a
    sibling if possible, else \textbf{fuse} with a sibling.
\end{itemize}
At the leaf, simply delete the key. Internal-node remove: replace
with in-order predecessor / successor from the appropriate subtree,
then remove that leaf key.

\subsection*{Insert vs.\ remove primitives}
\begin{tabular}{@{}l|l@{}}
\textbf{direction} & \textbf{primitive}\\
\hline
grow, make room    & split (node $\to$ two + median up)\\
shrink, share      & rotate (neighbor lends one)\\
shrink, collapse   & fusion (merge with neighbor)
\end{tabular}

\subsection*{Cost}
All three operations: $O(\log_m n)$ node reads; within a node
$O(\log m)$ search or $O(m)$ insert shift. If $m$ matches the disk
block, height is tiny ($\le 5$--6 for billions of keys).

\subsection*{B-tree vs.\ B+ tree}
\textbf{B+} keeps all data in leaves; internal nodes are pure
routing (just keys). Leaves form a linked list. Range queries are
then a single traversal along the leaf list. Every production
database (MySQL InnoDB, PostgreSQL, SQLite, LevelDB) ships a B+.

\subsection*{Top gotchas}
\begin{itemize}
    \item Using the wrong \emph{order} definition -- ``order $m$''
    means $m$ children max, $m-1$ keys max. Stay consistent.
    \item Forgetting the root is the only node allowed to have
    $< \lceil m/2 \rceil - 1$ keys.
    \item Splitting post-insertion instead of preemptively -- doable
    but forces a second pass.
    \item Remove without rotating/fusing first -- leaves a node with
    too few keys; violates the invariant.
    \item Using B-tree where a hash index would do. B-trees earn
    their keep on range queries and sorted iteration.
\end{itemize}

\end{multicols}
\end{document}
