\documentclass[11pt]{article}
\input{../../notes-style.tex}

\title{CS 300 -- Chapter 13 Notes\\\large Additional Material}
\author{Jose de Lima}
\date{\today}

\begin{document}
\maketitle

\begin{ideabox}
Chapter 13 is a grab-bag: three sorting algorithms (one pedagogical, one selection-only, one non-comparison) followed by two linked-list variants (the list-container idiom and the circular list). Each topic is self-contained; there's no unifying narrative. The common thread is ``useful ideas that didn't fit anywhere else.''
\end{ideabox}

\section{13.1 Bubble Sort}

\subsection*{The algorithm}

\begin{defnbox}
\textbf{Bubble sort.} Repeatedly walk the array, comparing adjacent pairs and swapping if out of order. After each pass, the largest remaining unsorted element ``bubbles'' to the end.
\end{defnbox}

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
void bubbleSort(std::vector<int>& a) {
    int n = a.size();
    for (int i = 0; i < n - 1; ++i) {
        for (int j = 0; j < n - i - 1; ++j) {
            if (a[j] > a[j + 1])
                std::swap(a[j], a[j + 1]);
        }
    }
}
\end{lstlisting}

The inner loop shrinks each pass because the last $i$ elements are already in their final position.

\subsection*{Cost}

\begin{itemize}[leftmargin=*]
    \item Best case (already sorted): still $O(n^2)$ comparisons in the plain version. An optimization: short-circuit when an entire pass made no swaps, giving $O(n)$ best case.
    \item Average: $O(n^2)$.
    \item Worst case (reverse sorted): $O(n^2)$ comparisons, $O(n^2)$ swaps.
    \item Space: $O(1)$.
    \item Stable: yes (equal elements never swap past each other).
\end{itemize}

\subsection*{Why you'll never use bubble sort}

\begin{warnbox}
Bubble sort is a \textbf{teaching tool}, nothing more. For any real data: \texttt{std::sort} (introsort), \texttt{std::stable\_sort} (merge sort), or a radix/bucket sort for integers. Bubble sort's $O(n^2)$ worst case on large inputs is not just slow -- it's catastrophically slow. $n = 10^6$ is $10^{12}$ operations, roughly an hour of CPU time. A real sort does the same job in milliseconds.
\end{warnbox}

\begin{notebox}
The one defense of bubble sort: its best case with the short-circuit optimization is $O(n)$, matching insertion sort. For nearly-sorted data, it's not terrible. But insertion sort is better on the same workload and easier to reason about, so even that niche has a better answer.
\end{notebox}

\subsection*{What bubble sort is good for}

\begin{itemize}[leftmargin=*]
    \item Teaching the concept of an in-place comparison sort.
    \item Interview questions about algorithmic complexity.
    \item Tiny inputs ($n \leq 20$) where the constant factor is irrelevant.
    \item Hardware sorting networks, where adjacent-pair compare-and-swap is the primitive operation.
\end{itemize}

\section{13.2 Quickselect}

\subsection*{The problem}

\begin{defnbox}
\textbf{$k$-th smallest element.} Given an array and an integer $k$, return the $k$-th smallest value. For $k=0$, the minimum; for $k = n-1$, the maximum; for $k = n/2$, the median.
\end{defnbox}

The naive solution: sort, return \texttt{a[k]}. That's $O(n \log n)$. Quickselect does it in $O(n)$ average time \emph{without} fully sorting.

\subsection*{The insight}

Quickselect borrows Quicksort's partition routine -- the one that picks a pivot, puts smaller elements to the left, larger to the right, and returns the pivot's final position.

\begin{ideabox}
\textbf{Quicksort recurses on both sides; quickselect recurses on only the side that contains the $k$-th element.} That cuts the total work from $O(n \log n)$ to $O(n)$ on average, because the unused half of each partition is discarded.
\end{ideabox}

\subsection*{Algorithm}

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
int partition(std::vector<int>& a, int lo, int hi) {
    int pivot = a[(lo + hi) / 2];
    int i = lo - 1, j = hi + 1;
    while (true) {
        do { ++i; } while (a[i] < pivot);
        do { --j; } while (a[j] > pivot);
        if (i >= j) return j;
        std::swap(a[i], a[j]);
    }
}

int quickselect(std::vector<int>& a, int lo, int hi, int k) {
    if (lo >= hi) return a[lo];
    int p = partition(a, lo, hi);
    if (k <= p) return quickselect(a, lo, p, k);
    return quickselect(a, p + 1, hi, k);
}
\end{lstlisting}

\subsection*{Cost}

\begin{itemize}[leftmargin=*]
    \item \textbf{Average}: $O(n)$. Partition costs $n$; the next recursion is on $n/2$; then $n/4$; \ldots total $n + n/2 + n/4 + \ldots = O(n)$ by geometric sum.
    \item \textbf{Worst}: $O(n^2)$, same as naive quicksort, when pivots are consistently bad.
    \item \textbf{Median-of-medians} gives $O(n)$ worst-case selection, but the constant is large; in practice quickselect with a random or median-of-three pivot is faster.
\end{itemize}

\subsection*{C++ standard library}

You don't have to write quickselect yourself -- C++ gives it to you:

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
#include <algorithm>
// Rearranges [begin, end) so nth is the element that would be at
// position n in a fully sorted range. Everything before nth is <=,
// everything after is >=. Average O(n).
std::nth_element(a.begin(), a.begin() + k, a.end());
int kthSmallest = a[k];
\end{lstlisting}

\begin{ideabox}
\texttt{std::nth\_element} is the correct answer to ``what's the median / top-$k$ / $k$-th percentile'' in any interview or real codebase. It's linear time, in place, and a single line of code. Reach for it before thinking harder.
\end{ideabox}

\subsection*{Applications}

\begin{itemize}[leftmargin=*]
    \item \textbf{Median / percentile computation.}
    \item \textbf{Top-$k$ queries.} After \texttt{nth\_element}, the first $k$ entries are the top-$k$ (unsorted). If you need them sorted, sort just that prefix.
    \item \textbf{Machine learning}: quickselect-based k-nearest-neighbor scoring.
    \item \textbf{Streaming statistics}: approximate quantiles via repeated quickselect.
\end{itemize}

\section{13.3 Bucket Sort}

\subsection*{The idea}

\begin{defnbox}
\textbf{Bucket sort.} Distribute values into $K$ buckets based on a range partition, sort each bucket with another algorithm, concatenate. $O(n)$ average when values are uniformly distributed.
\end{defnbox}

This is a \textbf{non-comparison sort} -- it exploits numerical structure of the values to beat the $\Omega(n \log n)$ comparison-sort lower bound.

\subsection*{Algorithm}

Given non-negative numbers up to some max value $M$:

\begin{enumerate}[leftmargin=*]
    \item Create $K$ buckets. Bucket $i$ holds values in $[i \cdot M/K, (i+1) \cdot M/K)$.
    \item For each element $x$, compute \texttt{bucket = floor(x * K / (M + 1))} and append $x$ to that bucket.
    \item Sort each bucket (insertion sort or any other; each bucket is small).
    \item Concatenate buckets in order.
\end{enumerate}

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
void bucketSort(std::vector<int>& a, int K) {
    if (a.empty()) return;
    int maxV = *std::max_element(a.begin(), a.end());
    std::vector<std::vector<int>> buckets(K);

    for (int x : a) {
        int idx = (long long)x * K / (maxV + 1);
        buckets[idx].push_back(x);
    }

    a.clear();
    for (auto& b : buckets) {
        std::sort(b.begin(), b.end());
        a.insert(a.end(), b.begin(), b.end());
    }
}
\end{lstlisting}

\subsection*{Cost}

Let $n$ be the array size and $K$ the bucket count.
\begin{itemize}[leftmargin=*]
    \item \textbf{Distribution pass}: $O(n)$.
    \item \textbf{Per-bucket sort}: if values are uniform, each bucket has $\approx n/K$ elements. Sorting each with $O(m^2)$ insertion sort gives total $K \cdot (n/K)^2 = n^2/K$. With $K = n$, this is $O(n)$.
    \item \textbf{Concatenate}: $O(n + K)$.
    \item \textbf{Total (uniform distribution, $K = n$)}: $O(n)$.
    \item \textbf{Worst case (all elements in one bucket)}: $O(n^2)$.
\end{itemize}

\begin{warnbox}
Bucket sort's $O(n)$ is \emph{contingent on distribution}. If all your values cluster in a narrow range (e.g., 100 random integers between 1 and $10^9$), bucket sort degenerates: 99 empty buckets, one bucket with everything. You're paying for $K$ buckets to do nothing. Use bucket sort only when you know the data is roughly uniform over a known range.
\end{warnbox}

\subsection*{Related algorithms}

\begin{itemize}[leftmargin=*]
    \item \textbf{Counting sort}: degenerate bucket sort where each bucket holds exactly one value. $O(n + M)$ time, $O(M)$ space. Perfect for small-range integers.
    \item \textbf{Radix sort}: iterate bucket sort on each digit. $O(n \cdot d)$ for $d$-digit numbers. Used in database sorts, GPU sorts, and networking.
    \item \textbf{Pigeonhole sort}: counting sort by another name, when value range $=$ number of elements.
\end{itemize}

\begin{ideabox}
\textbf{The $n \log n$ barrier only applies to comparison sorts.} Any sort that uses numeric / string structure of its inputs -- bucket, radix, counting -- can beat it for suitable data. Production string sorting (timsort in Python, dpqsort in Java) is comparison-based; production integer sorting (std::sort in libstdc++ at large sizes) can fall through to radix for speed.
\end{ideabox}

\section{13.4 List Data Structure}

\subsection*{The idiom}

A linked list can be represented in two ways:

\begin{itemize}[leftmargin=*]
    \item \textbf{Just a head pointer} (and maybe a tail pointer), held in local variables or as member fields.
    \item \textbf{A wrapper container} -- a \texttt{List} struct that owns \texttt{head}, \texttt{tail}, and maybe \texttt{size}. Functions take a \texttt{List*} instead of a raw head pointer.
\end{itemize}

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
// Approach 1: just nodes.
struct Node { int data; Node* next; };
Node* head = nullptr;
Node* tail = nullptr;
int size = 0;

// Approach 2: list wrapper.
struct List {
    Node* head = nullptr;
    Node* tail = nullptr;
    int size = 0;
};

void listPushBack(List* lst, int data) {
    Node* n = new Node{data, nullptr};
    if (!lst->head) lst->head = n;
    else            lst->tail->next = n;
    lst->tail = n;
    ++lst->size;
}
\end{lstlisting}

\subsection*{Why wrap?}

\begin{ideabox}
The wrapper exists so that list-manipulating functions take \emph{one} parameter instead of three (head, tail, size). More importantly, operations like ``insert at tail'' or ``get length'' become $O(1)$ because the struct already knows what the naked-node version would have to recompute.
\end{ideabox}

\begin{center}
\begin{tabular}{lll}
\hline
Operation & Head-only list & Wrapped list \\
\hline
Insert at head & $O(1)$ & $O(1)$ \\
Insert at tail & $O(n)$ (walk to end) & $O(1)$ (tail ptr) \\
Get size & $O(n)$ or maintained externally & $O(1)$ (cached) \\
Pass to function & 3 args (head, tail, size) & 1 arg (\texttt{List*}) \\
\hline
\end{tabular}
\end{center}

\subsection*{Why not just use \texttt{std::list}?}

\begin{warnbox}
In C++, \texttt{std::list} is the wrapper-style doubly-linked list and you should use it unless you have a specific reason not to. Writing your own list is a pedagogical exercise; in production, \texttt{std::list} (or \texttt{std::forward\_list} for singly-linked) is tested, debuggers know how to render it, and you don't leak memory on exception. The ``list data structure'' the textbook describes is reinventing \texttt{std::list}.
\end{warnbox}

\subsection*{When linked lists are the wrong answer}

Even with a nice wrapper, linked lists are frequently worse than \texttt{std::vector} (dynamic array):

\begin{itemize}[leftmargin=*]
    \item Cache locality: array elements are contiguous, list nodes scattered across the heap.
    \item Memory overhead: each node has a \texttt{next} pointer, often doubling memory vs.\ raw values.
    \item Random access: $O(n)$ for list, $O(1)$ for array.
    \item Iteration: much slower on a list due to cache misses on every \texttt{next}.
\end{itemize}

Use a linked list when you need cheap insertion/deletion at interior points with existing iterators, or when you're building something that requires pointer stability (e.g., an LRU cache with external references). Otherwise, default to \texttt{std::vector}.

\section{13.5 Circular Lists}

\subsection*{The definition}

\begin{defnbox}
\textbf{Circular linked list.} A linked list where the tail's \texttt{next} pointer points back to the head, not \texttt{nullptr}. Traversal wraps forever unless explicitly bounded.
\end{defnbox}

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
// Singly-linked circular
struct Node { int data; Node* next; };
// Last node's next points back to head.

// Doubly-linked circular (most common in practice)
struct DNode { int data; DNode* prev; DNode* next; };
// Last node's next -> head; head's prev -> last node.
\end{lstlisting}

\subsection*{Traversal}

The key trick: stop when you come back to the start, not when you hit \texttt{nullptr}.

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
void traverseCircular(Node* head) {
    if (!head) return;
    Node* cur = head;
    do {
        visit(cur);
        cur = cur->next;
    } while (cur != head);
}
\end{lstlisting}

\begin{warnbox}
\textbf{Use do-while, not while.} A regular \texttt{while (cur != head)} skips the head on the first iteration because we start at head and the condition fails immediately. The do-while runs the body first, then checks the condition.
\end{warnbox}

\subsection*{Why circular lists exist}

\begin{itemize}[leftmargin=*]
    \item \textbf{Round-robin scheduling.} OS schedulers cycle through ready processes; each yield pops one and reappends it at the end. A circular list gives O(1) ``next process'' and O(1) append.
    \item \textbf{Music players on ``repeat all.''} Queue is a circular list of tracks.
    \item \textbf{Ring buffers} (logically circular, though usually array-based for cache locality).
    \item \textbf{Josephus problem} and other classic cyclic-elimination algorithms -- naturally circular.
    \item \textbf{Token-passing networks.} Token Ring Ethernet was literally a circular list of nodes; each station had one successor.
    \item \textbf{Doubly-linked circular lists with a sentinel} give branch-free insert/delete at any position.
\end{itemize}

\subsection*{Sentinel trick for doubly-linked circular lists}

A \emph{sentinel} node -- a dummy that's never ``real'' data but completes the circle -- removes the ``is this node the head / tail'' special cases:

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
struct DNode { int data; DNode* prev; DNode* next; };

struct CircularList {
    DNode sentinel;   // sentinel.next = head, sentinel.prev = tail
    CircularList() { sentinel.prev = sentinel.next = &sentinel; }
};

// Insert n before 'where' -- no null checks needed.
void insertBefore(DNode* where, DNode* n) {
    n->next = where;
    n->prev = where->prev;
    where->prev->next = n;
    where->prev = n;
}

// Remove n -- no null checks needed.
void remove(DNode* n) {
    n->prev->next = n->next;
    n->next->prev = n->prev;
}
\end{lstlisting}

\begin{ideabox}
The sentinel technique eliminates all ``empty list'' and ``at the head'' edge cases. Insert/remove are six pointer writes with no branching. The Linux kernel's \texttt{list\_head} uses exactly this structure, and the design has been copied into countless production codebases -- it's one of the most elegant tricks in systems programming.
\end{ideabox}

\subsection*{Traversal in reverse (doubly-linked)}

Mirror of the forward case -- walk \texttt{prev} pointers from any starting point until you return:

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single]
void traverseReverseCircular(DNode* tail) {
    if (!tail) return;
    DNode* cur = tail;
    do {
        visit(cur);
        cur = cur->prev;
    } while (cur != tail);
}
\end{lstlisting}

\begin{ideabox}
\textbf{Chapter takeaway.} Bubble sort is useful as a pedagogical example of why most sorting algorithms are $O(n \log n)$. Quickselect solves a narrower problem (``find the $k$-th element'') faster than sorting and is what \texttt{std::nth\_element} uses internally. Bucket sort shows that when the $\Omega(n \log n)$ comparison barrier doesn't apply, you can sort in linear time. The list-container idiom is about encapsulating bookkeeping; circular lists are about cyclic problem domains. Nothing in this chapter changes how you think; each piece just adds one more specialized tool to the bag.
\end{ideabox}

\end{document}
