\documentclass[10pt]{article}
\input{../../notes-style.tex}

\usepackage{multicol}
\usepackage{array}

\title{Ch.~4 Cheat Sheet\\\large Lists, Stacks, Queues, Deques}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-2em}

\begin{multicols}{2}

\subsection*{List ADT operations}
\texttt{size()}, \texttt{empty()}, \texttt{front()}, \texttt{back()},
\texttt{append(x)}, \texttt{prepend(x)}, \texttt{insertAfter(cur, x)},
\texttt{remove(cur)}, \texttt{search(x)}, traversal.

\subsection*{SLL node}
\begin{lstlisting}[language=C++]
struct Node {
    int value;
    Node* next;       // nullptr = end
};
struct SLL { Node* head = nullptr;
             Node* tail = nullptr;
             size_t sz  = 0; };
\end{lstlisting}

\subsection*{DLL node}
\begin{lstlisting}[language=C++]
struct Node {
    int value;
    Node* next;
    Node* prev;
};
\end{lstlisting}
\textbf{Invariant:} for all nodes except head/tail,
\texttt{n->next->prev == n} and \texttt{n->prev->next == n}.

\subsection*{Append + prepend (SLL, with tail)}
\textbf{Append:} new node, \texttt{tail->next = new}, \texttt{tail = new}.
If empty, also set \texttt{head}.\\
\textbf{Prepend:} new node, \texttt{new->next = head}, \texttt{head = new}.
If empty, also set \texttt{tail}.\\
Cost: both $O(1)$.

\subsection*{Insert-after, remove (SLL)}
\textbf{insertAfter(cur, x):} three cases (empty, append, middle). Always
$O(1)$ once you have \texttt{cur}.\\
\textbf{remove} needs the \emph{predecessor} pointer -- if you only have
\texttt{cur}, you walk from head to find prev $\to$ $O(n)$.\\
This is why DLL exists.

\subsection*{DLL advantages over SLL}
\begin{itemize}
    \item \texttt{remove(cur)} in $O(1)$ when you have a pointer.
    \item \texttt{insertBefore(cur, x)} is as cheap as \texttt{insertAfter}.
    \item Reverse traversal is a natural \texttt{for (n=tail; n; n=n->prev)}.
\end{itemize}
Cost: two pointers per node (vs.\ SLL's one).

\subsection*{Sentinel (dummy) nodes}
Fixed head \emph{and} fixed tail node that hold no data. Every real node
is between them. Collapses the three cases (empty / first / last /
middle) into one uniform ``splice between prev and next'' code path.
\texttt{std::list} uses a circular sentinel.

\columnbreak

\subsection*{Linked-list search}
Linear: walk from head until match or \texttt{nullptr}.
Cost: $O(n)$. Cache-hostile: non-contiguous pointer chasing.

\subsection*{Stack ADT}
LIFO. \texttt{push(x)}, \texttt{pop()}, \texttt{top()}, \texttt{empty()}.
Implement with SLL (push/pop at head) or array/vector (push/pop at
back). Both $O(1)$. \texttt{std::stack} $\to$ defaults to
\texttt{std::deque}.

\subsection*{Queue ADT}
FIFO. \texttt{enqueue(x)}, \texttt{dequeue()}, \texttt{front()}.
SLL with both head and tail: enqueue at tail, dequeue at head. Array
alone is bad (dequeue from front is $O(n)$); use a ring buffer or
\texttt{std::deque}. \texttt{std::queue} $\to$ \texttt{std::deque}.

\subsection*{Deque ADT}
Push/pop at both ends. DLL is natural. Or block-based dynamic array
(\texttt{std::deque}).

\subsection*{Decision matrix}
\begin{tabular}{@{}l|cccc@{}}
\textbf{Op} & \textbf{arr} & \textbf{SLL} & \textbf{DLL} & \textbf{sDLL}\\
\hline
\texttt{v[i]}         & $1$ & $n$ & $n$ & $n$\\
append                & $1^*$ & $1$ & $1$ & $1$\\
prepend               & $n$ & $1$ & $1$ & $1$\\
insertAfter           & $n$ & $1$ & $1$ & $1$\\
insertBefore          & $n$ & $n$ & $1$ & $1$\\
remove (have ptr)     & $n$ & $n$ & $1$ & $1$\\
reverse traverse      & $n$ & $n$ & $n$ & $n$\\
bytes/elem            & lo  & + 1 ptr & + 2 ptrs & + 2 ptrs\\
cache-friendly        & yes & no  & no  & no
\end{tabular}

\subsection*{When to pick each}
\begin{itemize}
    \item \textbf{Array / vector.} Default. Random access. Dense
    memory.
    \item \textbf{SLL.} Rarely the right answer. Pick only if memory
    per element matters and all ops are at head.
    \item \textbf{DLL.} Mid-list inserts/removes with a held pointer,
    bidirectional traversal.
    \item \textbf{Sentinel DLL.} Same as DLL, with cleaner code. What
    \texttt{std::list} is.
\end{itemize}

\subsection*{Top gotchas}
\begin{itemize}
    \item Dereferencing \texttt{cur->next} after deletion (use-after-free).
    \item Forgetting to update \texttt{tail} on head operations.
    \item Missing the empty-list special case in \texttt{remove}.
    \item Assuming linked lists beat vectors on middle-insert (usually
    false under $10^4$).
    \item On a DLL, updating \texttt{next} without the twin
    \texttt{prev} update.
\end{itemize}

\end{multicols}
\end{document}
