\documentclass[10pt]{article}
\input{../../notes-style.tex}

\usepackage{multicol}
\usepackage{array}

\title{Ch.~12 Notes\\\large Sets}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-2em}

\begin{multicols}{2}

\subsection*{The Set ADT}
An unordered collection of \textbf{distinct} elements. Core ops:
\begin{itemize}
    \item \texttt{contains(x)} -- membership.
    \item \texttt{add(x)} -- insert; no-op if already present.
    \item \texttt{remove(x)} -- delete; no-op if absent.
\end{itemize}
Derived: \texttt{size}, \texttt{empty}, iteration. Everything else
(union, intersection, subset, filter, map) is built from these.

\subsection*{Keys vs.\ elements}
The set stores elements $x$ compared by some \emph{key} -- often the
element itself. When elements are complex (structs), you pass a
comparator (ordered set) or a hash + equality (unordered).

\subsection*{Subset testing}
$A \subseteq B$ iff $\forall x \in A : x \in B$. Iterate the smaller
side, probe the larger. $O(|A|)$ for hashed $B$, $O(|A| \log |B|)$
for tree-based.

\subsection*{Set operations}
\[
A \cup B,\ A \cap B,\ A \setminus B.
\]
Union / intersection commute; difference does not.

\begin{lstlisting}[language=C++]
// Hashed: iterate one, probe the other
for (auto& x : A)
    if (B.contains(x)) out.insert(x);

// Sorted (std::set): use STL merge-style
std::set_intersection(
    A.begin(), A.end(),
    B.begin(), B.end(),
    std::inserter(out, out.end()));
\end{lstlisting}

\subsection*{Always iterate the smaller set}
Intersection cost drops to $O(\min(|A|,|B|))$.
Single biggest micro-optimization in set-heavy code.

\subsection*{Cost table}
\begin{tabular}{@{}l|ll@{}}
                  & \textbf{hashed}        & \textbf{sorted (tree)}\\
\hline
contains          & $O(1)^*$               & $O(\log n)$\\
add / remove      & $O(1)^*$               & $O(\log n)$\\
union / inters.   & $O(|A| + |B|)$         & $O(|A| + |B|)$\\
sorted iter       & $O(n \log n)$          & $O(n)$\\
\texttt{lower\_bound} & --                 & $O(\log n)$
\end{tabular}
$^*$ expected.

\columnbreak

\subsection*{Static vs.\ dynamic}
\textbf{Static set}: fixed membership after construction; supports
read-only + new-producing ops. Allows compact encodings (sorted
array, bitset, perfect hash, Bloom filter).
\textbf{Dynamic}: supports \texttt{add} / \texttt{remove}. Pay more
per element for that flexibility.

\begin{tabular}{@{}l|l|l@{}}
\textbf{need}       & \textbf{static}       & \textbf{dynamic}\\
\hline
build once, query  & yes                   & overkill\\
small memory       & yes                   & harder\\
mutate at runtime  & no                    & yes\\
tuned lookup       & perfect hash, bitset  & hash / tree
\end{tabular}

\subsection*{Containers you'll actually use}
\begin{tabular}{@{}l|l|l@{}}
\textbf{container}               & \textbf{order}  & \textbf{impl.}\\
\hline
\texttt{std::set}                & sorted          & red-black tree\\
\texttt{std::multiset}           & sorted, dups    & red-black tree\\
\texttt{std::unordered\_set}     & hash order      & open-addr.\ hash\\
Python \texttt{set}              & hash            & open-addr.\ hash\\
Python \texttt{frozenset}        & hash, immutable & open-addr.\ hash\\
Rust \texttt{HashSet} / \texttt{BTreeSet} & -- / sorted & hash / B-tree
\end{tabular}

\subsection*{Picking the implementation}
\begin{itemize}
    \item Just need $O(1)$ membership, order doesn't matter:
    \texttt{unordered\_set}.
    \item Need sorted iteration, range queries, or
    \texttt{lower\_bound}: \texttt{set}.
    \item Fixed membership, lookup-only: sorted array +
    \texttt{std::binary\_search}, bitset, or perfect hash.
    \item Extremely tight memory, ``probably in set'' OK:
    Bloom filter (approximate).
\end{itemize}

\subsection*{Common patterns}
\begin{itemize}
    \item \textbf{Dedup}: \texttt{std::set<int> s(v.begin(), v.end())}
    or \texttt{sort + unique + erase}.
    \item \textbf{Seen-set} during traversal: avoid cycles in BFS / DFS.
    \item \textbf{Set as filter}: keep only items whose key is in a
    whitelist / not in a blacklist.
\end{itemize}

\subsection*{Top gotchas}
\begin{itemize}
    \item Relying on iteration order in \texttt{unordered\_set} --
    implementation-defined and libc-version-dependent.
    \item Mutating elements in a sorted set in a way that changes
    their sort key -- invalidates the invariant silently.
    \item Using \texttt{set\_intersection} / \texttt{set\_union} from
    \texttt{<algorithm>} on \emph{unsorted} ranges -- requires
    sorted input.
    \item Writing $A \setminus B$ as $A - B$ in math then forgetting
    which STL algorithm corresponds
    (\texttt{set\_difference}).
    \item Assuming ``const'' means truly immutable in C++ -- someone
    can \texttt{const\_cast} it away.
\end{itemize}

\end{multicols}
\end{document}
